A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-21 analysis/paper/paper.Rmd

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. The details about archaeological networks provided in Supplementary Materials. This paper will use a novel Bayesian approach on exponential random graph model to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

```{r site-map, figure-KWL-map, fig.cap = "The location of Kiwulan and other places with the European forts in northern Taiwan mentioned in the text (map modified from Wang and Marwick, 2020). Map data from naturalearthdata.com."}

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD). The settlement experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

<!-- chronology -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are eight burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. We reconstructed social relations between individuals by linking them according to similar prestige goods [cf. @Coward2013, pp. 252].Therefore, we built networks where burials represent actors (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts also support these items as prestige goods [@Borao2009; @LiandWu2006].

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of good between burials. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials. Gold-foil beads are in levels of high (>3), upper-middle (3), lower-middle (2), and low (1); carnelian beads are in levels of high (>6), upper-middle (4-6), lower-middle (3), and low (1-2); glass beads are in levels of high (>6), upper-middle (3-6), lower-middle (2), and low (1). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>28, about 10 percent), above average (12-28), and below average (<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship. The networks constructed based on this principles show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:figure-network-diagrams)).

>>>>

<!--burial study- network analysis-->
Burial analysis by archaeologists is an approach to understand past social structures through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identities of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors that structure the material configuration of burials, can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theory, the recent development of social network analysis provides many new tools to visualize and analyze relationships in archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. A network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data. The details about archaeological networks provided in Supplementary Online Materials. This paper will use a novel Bayesian approach on exponential random graph model for networks to investigate the formation of relationship between burials in northeastern Taiwan to explore the indirect impact of foreign contacts on social organization.

@@

```{r figure-KWL-map, fig.cap = "The location of Kiwulan and the European forts at Heping dao and Tamsui in northern Taiwan (modified from Wang and Marwick, 2020). Map data from naturalearthdata.com."}

@@

<!-- archaeological setting -->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD). The settlement experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and _in-situ_ wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

<!-- chronology -->
To compare burial networks, we assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`). Our assignments are based on an established fine-grained chronology that was reexamined and cross-validated by diagnostic materials, stratigraphic data, depth, and radiocarbon ages [@LWandBM2020ornament; @Chen2007]. We excluded burials from the Chinese phase (n = `r chi`) due to the smaller sample size. There are eight burials heavily disturbed by modern construction that are also excluded because we cannot determine their chronology. We reconstructed social relations between individuals by linking them according to similar prestige goods [cf. @Coward2013, pp. 252]. Therefore, we built networks where burials represent actors (nodes in the network) that are linked when they have the same prestige goods in common. The prestige goods we identified include agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments. Previous studies indicate that these items are considered as high-value across different archaeological contexts [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011]. Historical accounts also support these items as prestige goods [@Borao2009; @LiandWu2006].

@@

<!-- burial data and the definition of tie formation  --> 
We used burial goods in common to construct ties for representing a flow of good between burials. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials (details in Supplementary Online Materials). For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>28, about 10 percent), above average (12-28), and below average (<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship. The networks constructed based on this principles show that the network after the European presence has higher node degrees in general (Figure \@ref(fig:figure-network-diagrams)).
